---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# plotor

<!-- badges: start -->

<!-- badges: end -->

The goal of plotor is to ...

## Installation

You can install the development version of plotor from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("craig-parylo/plotor")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, message=FALSE}
library(plotor)
library(dplyr)
library(datasets)
library(tidyr)
library(stats)
library(broom)
library(forcats)
library(ggplot2)
```

In this example we will explore the likelihood of surviving the Titanic
disaster based on passenger economic status (class), sex, and age group.

Start with getting the data from the datasets package.

```{r}
df <- datasets::Titanic |> 
  as_tibble() |> 
  # convert counts to observations
  filter(n > 0) |>
  uncount(weights = n) |>
  # convert categorical variables to factors.
  # we specify an order for levels in Class and Survival, otherwise ordering
  # in descending order of frequency
  mutate(
    Class = Class |>
      fct(levels = c('1st', '2nd', '3rd', 'Crew')),
    Sex = Sex |>
      fct_infreq(),
    Age = Age |>
      fct_infreq(),
    Survived = Survived |>
      fct(levels = c('No', 'Yes'))
  )
```

We now have a tibble of data containing four columns:

-   `Survived` - our outcome variable describing whether the passenger
    survived `Yes` or died `No`,

-   `Class` - the passenger class, either `1st`, `2nd`, `3rd` or `Crew`,

-   `Sex` - the gender of the passenger, either `Male` or `Female`,

-   `Age` - whether the passenger was an `Adult` or `Child`.

We next conduct a logistic regression of survival (as a binary factor:
'yes' and 'no') against the characteristics of passenger class, sex and
age group. For this we use the Generalised Linear Model function (`glm`)
from the `stats` package, specifying:

-   the family as 'binomial', and

-   the formula as survival being a function of `Class`, `Sex` and
    `Age`.

```{r}
# conduct a logistic regression of survival against the other variables
lr <- glm(
  data = df, 
  family = 'binomial', 
  formula = Survived ~ Class + Sex + Age
  )
```

Finally, we can plot the Odds Ratio of survival using the `plot_or`
function.

```{r warning=FALSE, fig.width=9, fig.height=6, fig.format='svg'}
# using plot_or 
plot_or(glm_model_results = lr)
```
