---
title: "Using plotor"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using plotor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(plotor)
library(dplyr)
library(datasets)
library(tidyr)
library(stats)
library(broom)
library(forcats)
library(ggplot2)
```

`plotor` is designed to produce Odds-Ratio plots from a given logistic regression model as produced using the general linear model (glm) package.

## Installing plotor

`plotor` can be installed via GitHub using the `devtools` package:

```{r eval=FALSE}
devtools::install_github()
```

## Example 1 - using the Titanic survivors dataset

In this example we will explore the likelihood of surviving the Titanic disaster based on passenger economic status (class), sex, and age group.

Start with getting the data from the `datasets` package.

```{r}
df <- datasets::Titanic |> 
  as_tibble() |> 
  # convert counts to observations
  filter(n > 0) |>
  uncount(weights = n) |>
  # convert categorical variables to factors.
  # we specify an order for levels in Class and Survival, otherwise ordering
  # in descending order of frequency
  mutate(
    Class = Class |>
      fct(levels = c('1st', '2nd', '3rd', 'Crew')),
    Sex = Sex |>
      fct_infreq(),
    Age = Age |>
      fct_infreq(),
    Survived = Survived |>
      fct(levels = c('No', 'Yes'))
  )

# preview the data
df |> 
  glimpse()
```

We now have a tibble of data containing four columns:

-   `Survived` - our outcome variable describing whether the passenger survived `Yes` or died `No`,

-   `Class` - the passenger class, either `1st`, `2nd`, `3rd` or `Crew`,

-   `Sex` - the gender of the passenger, either `Male` or `Female`,

-   `Age` - whether the passenger was an `Adult` or `Child`.

We next conduct a logistic regression of survival (as a binary factor: 'yes' and 'no') against the characteristics of passenger class, sex and age group. For this we use the Generalised Linear Model function (`glm`) from the `stats` package, specifying:

-   the family as 'binomial', and

-   the formula as survival being a function of `Class`, `Sex` and `Age`.

```{r}
# conduct a logistic regression of survival against the other variables
lr <- glm(
  data = df,
  family = 'binomial',
  formula = Survived ~ Class + Sex + Age
)

# preview the model as a tidy table
lr |> 
  tidy() |> 
  glimpse()
```

Finally, we can plot the Odds Ratio of survival using the `plot_or` function.

```{r warning=FALSE, fig.width=9, fig.height=6, fig.format='svg'}
# using plot_or
plot_or(glm_model_results = lr)
```

### Change base font size

To increase the size of the font you can extend the returned plot using the `theme` function from `ggplot2`. Here we set the base size of all fonts in the plot to size 16.

```{r warning=FALSE, fig.width=9, fig.height=6, fig.format='svg'}
plot_or(glm_model_results = lr) + 
  theme(text = element_text(size = 16))
```

### Specify the breaks

Odds Ratio (OR) plots produced by `plotor` are displayed using a log10 x axis.

By default ten breaks are shown, however, this can be altered by extending the `scale_x_log10` function from `ggplot2`. Here we provide a manual list of breaks to use:

```{r warning=FALSE, message=FALSE, fig.width=9, fig.height=6, fig.format='svg'}
plot_or(glm_model_results = lr) + 
  scale_x_log10(breaks = c(0.1, 0.5, 1, 5, 10))
```

### Change the dot and whisker colours

There are three types of colours used for the dots and whiskers in the OR plot, depending on their category.

-   `Significant` refers to dots where their results indicate a significant finding because their 95% confidence intervals do not touch or cross the value 1 - the line of no effect.

-   `Comparator` refers to the level of a factor in the model against which the Odds Ratios are calculated.

-   `Not significant` refers to dots where their results do not indicate a significant finding because their confidence intervals touch or cross the line of no effect.

The colours for these points can be changed by extending the output using `scale_colour_manual` function from `ggplot2` with a named vector specifying colour values for the three types of colours:

```{r warning=FALSE, message=FALSE, fig.width=9, fig.height=6, fig.format='svg'}
plot_or(glm_model_results = lr) +
  scale_colour_manual(values = c(
    'Significant' = '#44bd32',
    'Comparator' = '#8c7ae6',
    'Not significant' = '#e84118')
  )
```

### Change the title, subtitle and add a caption

`plotor` uses the dependent variable as the title of the plot by default with a subtitle indicating this it is an Odds Ratio plot with a 95% confidence interval.

The plot can be customised with your own title, subtitle and add a caption by extending the `labs` function of `ggplot2`.

```{r warning=FALSE, message=FALSE, fig.width=9, fig.height=6, fig.format='svg'}
plot_or(glm_model_results = lr) +
  labs(
    title = 'Passenger survival from the Titanic disaster',
    subtitle = 'Odds Ratio of survival by Class, Age and Gender',
    caption = 'Data source: Dawson, Robert J. MacG. (1995), The ‘Unusual Episode’ Data Revisited. Journal of Statistics Education, 3. doi:10.1080/10691898.1995.11910499'
  )
```
