---
title: "Creating publication-ready labels for your outputs"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(plotor)
set.seed(123)
```

## Overview

By default, `plot_or()` and `table_or()` display variable names exactly as they appear in your model. Variable names like `age_group_35_44` or `smoking_status_yes` are **technical and not suitable for publication-ready outputs**.

The solution is to **attach human-readable labels to variables**. When you do, `plotor` automatically uses these labels in plots and tables instead of the raw variable names. This approach is **much cleaner than manually editing outputs** after generation.

# Example 1: Oesophageal Cancer Study

## The dataset

This [case-control study](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/esoph.html) examined oesophageal cancer in Ile-et-Vilaine, France. It contains:

| Variable | Description                             |
|----------|-----------------------------------------|
| `Group`  | Case (cancer) or Control (disease-free) |
| `agegp`  | Age group of participant                |
| `alcgp`  | Alcohol consumption (grams per day)     |
| `tobgp`  | Tobacco consumption (grams per day)     |

## Preparing the data

```{r}
#| label: ex1 - dataset wrangling

# prepare the dataset for modelling
df <- 
  datasets::esoph |> 
  # convert aggregated data to tidy observational data
  tidyr::pivot_longer(
    cols = c(ncases, ncontrols),
    names_to = 'Group',
    values_to = 'people'
  ) |> 
  tidyr::uncount(weights = people) |> 
  # prepare the variables
  dplyr::mutate(
    # convert the intervention group to a factor
    Group = Group |> 
      dplyr::recode_values(
        "ncases" ~ "Case",
        "ncontrols" ~ "Control"
      ) |> 
        factor(levels = c("Control", "Case")),
    # remove ordering from these predictors
    agegp = agegp |> factor(ordered = FALSE),
    alcgp = alcgp |> factor(ordered = FALSE),
    tobgp = tobgp |> factor(ordered = FALSE)
  )

# preview the data
df |> dplyr::glimpse()
```

## Without labels

```{r}
#| label: ex1 - model without labels
#| warning: false

m <- glm(
  data = df,
  family = "binomial",
  formula = Group ~ agegp + alcgp + tobgp
)

# plot the odds ratio with a customised title
plot_or(m)
```

Notice how the plot uses technical variable names like `alcgp` and `tobgp` which are not immediately clear to readers.

## Adding labels with `{labelled}`

To make your outputs more readable, attach descriptive labels to your variables before modelling.

First, ensure the package is installed.

``` r
install.packages("labelled")
library(labelled)
```

```{r}
#| label: ex1 - attach labels

# create a list that matches variables with user-friendly labels
var_labels <- list(
  agegp = "Age group",
  alcgp = "Alcohol consumption",
  tobgp = "Tobacco consumption",
  Group = "Likelihood of developing oesophageal cancer"
)

# apply these variables to our data
labelled::var_label(df) <- var_labels

# preview the data with labels applied
labelled::look_for(df)
```


### Forest plots with labels

Now fit the model using the labelled data:

```{r}
#| label: ex1 - model with labels
#| warning: false

m <- glm(
  data = df,
  family = "binomial",
  formula = Group ~ agegp + alcgp + tobgp
)

# plot the odds ratio with a customised title
plot_or(m)
```

The plot is now much more **reader-friendly**:

-   The outcome label ("Likelihood of developing oesophageal cancer") appears in the title

-   Predictor labels ("Age group", "Alcohol consumption", "Tobacco consumption") replace technical variable names

### Data tables with labels

`table_or()` also respects variable labels:

```{r}
#| label: ex1 - table with labels

table_or(m, output = "gt", assumption_checks = FALSE)
```

# Example 2: Post-endoscopic pancreatitis study

## The dataset

The `indo_rct` dataset contains details from **602 patients** in a randomised controlled study examining indomethacin vs placebo for preventing post-endoscopic pancreatitis.

```{r}
#| label: ex2 - dataset wrangling

# prepare the dataset for modelling
df <- medicaldata::indo_rct |> 
  tibble::as_tibble() |> 
  # clean up factor levels
  dplyr::mutate(
    rx = forcats::fct_recode(
      .f = rx, 
      Placebo = "0_placebo", Indomethacin = "1_indomethacin"),
    pep = forcats::fct_recode(
      .f = pep,
      No = "0_no", Yes = "1_yes"
    ),
    amp = forcats::fct_recode(
      .f = amp,
      No = "0_no", Yes = "1_yes"
    )
  )

# fit the model without labels
m <- stats::glm(
  formula = outcome ~  rx + pep + amp,
  family = "binomial",
  data = df
)

plot_or(m, assumption_checks = FALSE)
```

## Using `{Hmisc}` for labels

An alternative to {labelled} is the [{Hmisc}](https://hbiostat.org/r/hmisc/) package, which is particularly useful if you're already using other {Hmisc} functions like `describe()` for summary statistics.

Install it with:

``` r
install.packages("Hmisc")
library(Hmisc)
```

Attach labels using the `label()` function:

```{r}
#| label: ex2 - labelling with Hmisc

# label the variables
Hmisc::label(df$outcome) <- "Likelihood of post-ERCP pancreatitis"
Hmisc::label(df$rx) <- "Treatment arm"
Hmisc::label(df$pep) <- "Previous post-ERCP pancreatitis (PEP)"
Hmisc::label(df$amp) <- "Ampullectomy performed"

# fit the model with labelled data
m <- stats::glm(
  formula = outcome ~  rx + pep + amp,
  family = "binomial",
  data = df
)

# plot
plot_or(m, assumption_checks = FALSE)
```

This plot now clearly shows that **treatment with Indomethacin** has a protective effect against pancreatitis, whereas a **history of pancreatitis** and **ampullectomy** are both associated with increased risk.

The same labels also appear in covariate tables:

```{r}
#| label: ex2 - table with Hmisc labels
table_or(m, output = "gt", assumption_checks = FALSE)
```

# Best practices for labelling

## Be descriptive but concise

Keep labels clear and specific, but not verbose:

| **✓ Good**                       | **✗ Avoid**                           |
|----------------------------------|---------------------------------------|
| "Age Group (years)"              | "The age of the participant in years" |
| "Systolic Blood Pressure (mmHg)" | "BP_sys"                              |
| "Smoking Status"                 | "smoking_status_yes"                  |

## Include units where relevant

Always specify units in parentheses:

``` r
var_labels <- list(
  wt = "Weight (kg)",
  ht = "Height (cm)",
  bp_sys = "Systolic Blood Pressure (mmHg)"
)
```

## Use consistent formatting

Apply consistent capitalisation and punctuation across all labels:

``` r
# consistent approach to capitalisation
var_labels <- list(
  ag_gp = "Age Group",
  sm_status = "Smoking Status",
  ed_level = "Education Level"
)
```

### Label factor levels clearly

Make factor level labels explicit and unambiguous:

```{r}
df <-
  data.frame(
    education = sample(
      x = 1:3,
      size = 10,
      replace = TRUE
    ) |>
      factor(
        labels = c(
          "Primary school",
          "Secondary school",
          "University degree"
        )
      )
  )
```

## Preserving labels in your workflow

### R-native formats preserve labels

Labels are preserved when saving and reloading with `.Rds` or `.RData`:

``` r
# labels are preserved
saveRDS(df, "my_data.Rds")
df <- readRDS("my_data.Rds")
```

### CSV files lose labels

**Labels are lost** when reading from CSV or other text formats. To preserve labels, use one of these approaches:

1.  **Use R-native formats** like `.Rds` or `.RData`

2.  **Re-apply labels** after reading CSV files

3.  **Store labels separately** in a data dictionary and apply them in your analysis script

## See also

-   `vignette("table_or")` - formatting results tables and exporting with {gt}

-   `vignette("check_or")` - diagnostics and model validation

-   [labelled](https://larmarange.github.io/labelled/) - comprehensive labelling package

-   [haven](https://haven.tidyverse.org) - import labelled data from SPSS, Stata or SAS

-   [Hmisc](https://hbiostat.org/r/hmisc/) - labelling and statistical functions commonly used in epidemiology and biostatistics